Positional surprisal (1-3grams) in CHILDES corpora

### 2019-04-25 surprisal/childes_surprisal.Rmd
fixed childes surprisal trigrams

---
title: Positional entropy, conditional entropy and mutual information in English CHILDES corpora
author: Josef Klafka and Dan Yurovsky 
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
    number_sections: false
    theme: lumen
    toc_float: false
    code_folding: show 
---

@@

Get utterances
```{r get_utterances}
prov_utterances <- get_utterances(corpus = "Providence") #stem or gloss
```

```{r}
adult_utterances <- prov_utterances %>%
  filter(speaker_role == "Target_Child") %>% 
  mutate(length = str_count(gloss, pattern = "[ +]+") + 1) %>%
  mutate(utterance_id = 1:n()) %>%
  unnest_tokens(word, gloss) %>%
  group_by(utterance_id) %>%
  mutate(word_order = 1:n())
```

@@

bigram_surprisals_nonnull %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)

```{r trigrams}
adult_trigrams <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  group_by(lag_word2, lag_word1, word) %>%
  count() %>%
  filter(!is.na(lag_word1), !is.na(lag_word2)) %>%
  ungroup() %>%
  mutate(tri_joint_p = n / sum(n)) %>%
  select(-n) %>%
  left_join(adult_bigrams, by = c("lag_word2" = "lag_word", "lag_word1" = "word")) %>%
  mutate(tri_cond_p = tri_joint_p / joint_p) %>%
  select(-joint_p, -tri_joint_p)

trigram_surprisals_prep <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  left_join(adult_trigrams) %>% #add bigram and trigram conditionals
  left_join(adult_unigrams)

trigram_surprisals_nonnull <- trigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word1), # check if it's not the first word
                    ifelse(is.na(lag_word2), # check if not the second word
                           -log(p), # trigram if third or beyond
                           -log(cond_p)), # bigram if second word
                    -log(tri_cond_p))) %>% # unigram if first word
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

@@

```{r make plots for joint lab meeting}
df <- prov_utterances %>%
  extract_stem_tokens(role = "adult", min_length = 5, max_length = 9)

prov_child_tokens <- extract_stem_tokens("child", 5, 9, prov_utterances) 
prov_adult_tokens <- extract_stem_tokens("not child",5, 9, prov_utterances)

prov_child_entropies <-  map_df(c(5,7,9), ~unigram_entropy(.x, prov_child_tokens)) %>%
  mutate(language = "English", person = "Child")
prov_adult_entropies <-  map_df(c(5,7,9), ~unigram_entropy(.x, prov_adult_tokens)) %>%
  mutate(language = "English", person = "Adult")

prov_entropies <- prov_child_entropies %>%
  bind_rows(prov_adult_entropies)

prov_entropies %>% ggplot(aes(x = word_order, y = empirical_entropy, color = person, label = person)) + 
  facet_wrap(~ length) + 
  xlab("Word position") + 
  ylab("Positional entropy") + 
  geom_pointrange(size = .25, aes(ymin = ci_lower, ymax = ci_upper)) + 
  geom_line(size = .25) + 
  theme_classic(base_size = 10) + 
  scale_color_ptol() + 
  theme(legend.position = "nones") + 
  geom_dl(aes(label = person), method = "last.points") 
```
bootstrap entropy slopes
```{r slope bootstrapping}

#extract tokens to be used for slope bootstrapping
extract_stem_tokens <- function(role, target_length, utterances) {

  if (role == "child") {
    sub_utterances <- utterances %>%
      filter(speaker_role == "Target_Child") %>%
      mutate(length = str_count(stem, pattern = " +") + 1)

} else {
    sub_utterances <- utterances %>%
      filter(speaker_role != "Target_Child") %>%
      mutate(length = str_count(stem, pattern = " +") + 1)
}

  tokens <- sub_utterances %>%
    filter(length == target_length) %>%
    mutate(utterance_id = 1:n()) %>%
    unnest_tokens(word, stem) %>%
    group_by(utterance_id) %>%
    mutate(word_order = 1:n()) %>%
    group_by(word_order, word) %>%
    filter(word_order <= length)

  tokens %>%
    summarise(n = n()) %>%
    tidyboot(summary_function = function(x) x %>% 
               summarise(entropy = entropy(n, unit = "log2")),
             statistics_functions = function(x) x %>%
             summarise_at(vars(entropy), funs(ci_upper, ci_lower))) %>%
    mutate(role = role, length = target_length)
}

extract_gloss_tokens <- function(role, target_length, utterances) {

  if (role == "child") {
    sub_utterances <- utterances %>%
      filter(speaker_role == "Target_Child") %>%
      mutate(length = str_count(gloss, pattern = " +") + 1)

} else {
    sub_utterances <- utterances %>%
      filter(speaker_role != "Target_Child") %>%
      mutate(length = str_count(gloss, pattern = " +") + 1)
}

  tokens <- sub_utterances %>%
    filter(length == target_length) %>%
    mutate(utterance_id = 1:n()) %>%
    unnest_tokens(word, gloss) %>%
    group_by(utterance_id) %>%
    mutate(word_order = 1:n()) %>%
    group_by(word_order, word) %>%
    filter(word_order <= length)

  tokens %>%
    summarise(n = n()) %>%
    tidyboot(summary_function = function(x) x %>% 
               summarise(entropy = entropy(n, unit = "log2")),
             statistics_functions = function(x) x %>%
             summarise_at(vars(entropy), funs(ci_upper, ci_lower))) %>%
    mutate(role = role, length = target_length)
}

#compute the unigram entropy measure from the counts of each word in each position
unigram_entropy <- function(tokens) {

  tokens %>%
    group_by(word_order, word) %>%
    summarise(n = n()) %>%
    tidyboot(summary_function = function(x) x %>% 
               summarise(entropy = entropy(n, unit = "log2")),
             statistics_functions = function(x) x %>%
             summarise_at(vars(entropy), funs(ci_upper, ci_lower))) %>%
    mutate(role = role, length = sen_length)
}

start_slope <- function(df) {

 df %>%
    filter(word_order %in% c(1,2)) %>%
    sample_frac(1, replace = T) %>%
    group_by(length, word_order) %>%
    summarise(entropy = entropy(n, unit = "log2")) %>%
    spread(word_order, entropy) %>%
    ungroup() %>%
    mutate(diff = `2` - `1`) %>%
    summarise(diff = mean(diff)) %>%
    pull(diff)

}

second_slope <- function(df) {

 df %>%
    ungroup() %>%
    mutate(word_order %in% c(2, 3)) %>%
    filter(word_order > 0) %>%
    group_by(length, word_order) %>%
    sample_frac(1, replace = T) %>%
    group_by(length, word_order) %>%
    summarise(entropy = entropy(n, unit = "log2")) %>%
    spread(word_order, entropy) %>%
    ungroup() %>%
    mutate(diff = `3` - `2`) %>%
    summarise(diff = mean(diff)) %>%
    pull(diff)

}

#note this measurement only makes sense if utterance length is at least 6
mid_slope <- function(df) {

 df %>%
    ungroup() %>%
    mutate(word_order = if_else(word_order == length-2, 2,
                                if_else(word_order == 3, 1,
                                        0))) %>%
    filter(word_order > 0) %>%
    group_by(length, word_order) %>%
    sample_frac(1, replace = T) %>%
    group_by(length, word_order) %>%
    summarise(entropy = entropy(n, unit = "log2")) %>%
    spread(word_order, entropy) %>%
    ungroup() %>%
    mutate(diff = `2` - `1`) %>%
    summarise(diff = mean(diff)) %>%
    pull(diff)

}

penult_slope <- function(df) {

 df %>%
    ungroup() %>%
    mutate(word_order = if_else(word_order == length-1, 2,
                                if_else(word_order == length -2, 1,
                                        0))) %>%
    filter(word_order > 0) %>%
    group_by(length, word_order) %>%
    sample_frac(1, replace = T) %>%
    group_by(length, word_order) %>%
    summarise(entropy = entropy(n, unit = "log2")) %>%
    spread(word_order, entropy) %>%
    ungroup() %>%
    mutate(diff = `2` - `1`) %>%
    summarise(diff = mean(diff)) %>%
    pull(diff)

}

end_slope <- function(df) {

 df %>%
    ungroup() %>%
    mutate(word_order = if_else(word_order == length, 2,
                                if_else(word_order == length -1 , 1,
                                        0))) %>%
    filter(word_order > 0) %>%
    group_by(length, word_order) %>%
    sample_frac(1, replace = T) %>%
    group_by(length, word_order) %>%
    summarise(entropy = entropy(n, unit = "log2")) %>%
    spread(word_order, entropy) %>%
    ungroup() %>%
    mutate(diff = `2` - `1`) %>%
    summarise(diff = mean(diff)) %>%
    pull(diff)

}

get_slopes <- function(role, target_length, utterances, stemgloss) {

  if (stemgloss == "stem") {
    tokens <- extract_stem_tokens(role, target_length, utterances)  
  } else {
    tokens <- extract_tokens(role, target_length, target_length, utterances)
  }

  start_slopes <- replicate(1000, start_slope(tokens)) %>% 
    as_data_frame() %>%
    summarise(mean = mean(value), ci_upper = quantile(value, .975), 
              ci_lower = quantile(value, .025))

  second_slopes <- replicate(1000, second_slope(tokens)) %>% 
    as_data_frame() %>%
    summarise(mean = mean(value), ci_upper = quantile(value, .975), 
              ci_lower = quantile(value, .025))

  mid_slopes <- replicate(1000, mid_slope(tokens)) %>% 
    as_data_frame() %>%
    summarise(mean = mean(value), ci_upper = quantile(value, .975), 
              ci_lower = quantile(value, .025))

  penult_slopes <- replicate(1000, penult_slope(tokens)) %>% 
    as_data_frame() %>%
    summarise(mean = mean(value), ci_upper = quantile(value, .975), 
              ci_lower = quantile(value, .025))

  end_slopes <- replicate(1000, end_slope(tokens)) %>% 
    as_data_frame() %>%
    summarise(mean = mean(value), ci_upper = quantile(value, .975), 
              ci_lower = quantile(value, .025))

  rbind(start_slopes, second_slopes, mid_slopes, penult_slopes, end_slopes)
}
```

Compute Providence Corpus Estimates
```{r calc_pos_entropy}
tokens <- extract_stem_tokens("not child", 6, 30, prov_utterances)

get_counts <- function(role, sen_length, utterances) {

  if(role == "Target_Child") {
    sub_utterances <- utterances %>%
      filter(speaker_role == "Target_Child") %>%
      mutate(length = str_count(gloss, " ") + 1) # uses stems of words, not original words
  } else {
     sub_utterances <- utterances %>%
      filter(speaker_role != "Target_Child") %>%
      mutate(length = str_count(gloss, " ") + 1)
  }

  sub_utterances %>%
    filter(length == sen_length) %>%
    mutate(utterance_id = 1:n()) %>%
    unnest_tokens(word, gloss) %>%
    group_by(utterance_id) %>%
    mutate(word_order = 1:n()) %>%
    filter(word_order <= length)
}

get_mean_lengths <- function(role, sen_length, utterances) {

  counts <- get_counts(role, sen_length, utterances)

  lens <- counts %>% 
    group_by(word_order, word) %>% 
    mutate(word_len = str_length(word))

  lens$word_len %>% 
    aggregate(list(lens$word_order), mean)
}

get_entropies <- function(role, sen_length, utterances) {

  if(role == "Target_Child") {
    sub_utterances <- utterances %>%
      filter(speaker_role == "Target_Child") %>%
      mutate(length = str_count(gloss, " ") + 1) # uses stems of words, not original words
  } else {
     sub_utterances <- utterances %>%
      filter(speaker_role != "Target_Child") %>%
      mutate(length = str_count(gloss, " ") + 1)
  }

  tokens <- sub_utterances %>%
    filter(length == sen_length) %>%
    mutate(utterance_id = 1:n()) %>%
    unnest_tokens(word, gloss) %>%
    group_by(utterance_id) %>%
    mutate(word_order = 1:n()) %>%
    group_by(word_order, word) %>%
    filter(word_order <= length)

  tokens %>%
    summarise(n = n()) %>%
    tidyboot(summary_function = function(x) x %>% 
               summarise(entropy = entropy(n, unit = "log2")),
             statistics_functions = function(x) x %>%
             summarise_at(vars(entropy), funs(ci_upper, ci_lower))) %>%
    mutate(role = role, length = sen_length)
}

tokens %>%
    filter(word_order %in% c(1,2)) %>%
    dplyr::group_by(length, word_order) %>%
    dplyr::summarise(entropy = entropy(n, unit = "log2"), n = n()) %>%
    spread(word_order, entropy) %>%
    group_by(length) %>%
    mutate(diff = `2` - `1`) %>%
  ggplot(aes(x = length, y = diff)) +
  geom_point(aes(size = n)) +
  geom_smooth(se = F) +
  scale_x_log10() #log length stats

entropies <- map(2:10, ~get_entropies("not child", .x, brown_utterances)) %>%
  bind_rows()

ggplot(entropies, aes(x = word_order, y = empirical_entropy,
                      ymin = ci_lower, ymax = ci_upper)) +
  facet_wrap(~ length) + 
  geom_pointrange() +
  geom_smooth(se = F)

entropies_child <- map(2:10, ~ get_entropies("Target_Child", .x, prov_utterances)) %>%
  bind_rows()

ggplot(entropies_child, aes(x = word_order, y = empirical_entropy,
                      ymin = ci_lower, ymax = ci_upper)) +
  facet_wrap(~ length) + 
  geom_pointrange() +
  geom_smooth(se = F)

entropies_all <- bind_rows(entropies, entropies_child)

ggplot(entropies_all, aes(x = word_order, y = empirical_entropy,
                      ymin = ci_lower, ymax = ci_upper, color = role)) +
  facet_wrap(~ length) + 
  geom_pointrange(position = position_dodge(.25)) +
  geom_smooth(se = F) +
  ggtitle("Child and Adult Positional Entropy vs. Word Position (Providence corpus)")

English Mutual Information
```{r mutual_information}

#compute the pairwise Mutual Information measure between all pairs of word positions
all_pairwise_mi <- function(df) {
  max_len <- max(df$word_order)

  combs <- combn(1:max_len, 2, simplify = F)

  map(combs, ~position_mi(df, .x[1], .x[2])) %>%
    bind_rows()

}

#perform the actual Mutual Information computation
position_mi <- function(df, pos1, pos2) {

  mi <- df %>%
    filter(word_order %in% c(pos1, pos2)) %>%
    mutate(word_order = factor(word_order, labels = c("first", "second"))) %>%
    spread(word_order, word) %>%
    group_by(first, second) %>%
    summarise(n = n()) %>%
    spread(second, n, fill = 0) %>% 
    ungroup() %>%
    select(-first) %>%
    mi.empirical(., unit = "log2")

  data_frame(length = max(df$word_order), pos1 = pos1, pos2 = pos2, mi = mi)
}

counts <- map(2:6, ~get_counts("not child", .x, prov_utterances))

mi <- map(counts, all_pairwise_mi) %>% bind_rows()

ggplot(mi, aes(x = pos2, y = mi, color = as.factor(pos1))) + 
  facet_wrap(~length) + 
  geom_point() + 
  geom_line()

ggplot(mi, aes(x = pos1, y = mi, color = as.factor(pos2))) + 
  facet_wrap(~length) + 
  geom_point() + 
  geom_line()

English Conditional Entropy
```{r conditional_entropy}
#compute all Conditional Entropy from all one-way pairs of word positions
all_pairwise_ce <- function(df) {
  max_len <- max(df$word_order)

  perms <- permutations(max_len, 2, 1:max_len, repeats.allowed = F) 
  combs <-  setNames(split(perms, seq(nrow(perms))), rownames(perms))

  map(combs, ~conditional_entropy(df, .x[1], .x[2])) %>%
    bind_rows()

}

#perform the actual Conditional Entropy calculation
get_ce <- function(df, X, Y) { #Y is conditioning variable

  d <- df %>%
    filter(word_order %in% c(X, Y)) %>%
    mutate(word_order = factor(word_order, labels = c("first", "second"))) %>%
    spread(word_order, word)

  ent <- d$first %>% table() %>% as.vector() %>% entropy.empirical(unit="log2")

  mi <- d %>% 
    group_by(first, second) %>%
    summarise(n = n()) %>%
    spread(second, n, fill = 0) %>% 
    ungroup() %>%
    select(-first) %>%
    mi.empirical(., unit = "log2")

  data_frame(length = max(df$word_order), X = X, Y = Y, mi = mi, ent = ent, ce = ent - mi)
}

counts <- map(2:8, ~get_counts("not child", .x, prov_utterances))

ce <- map(counts, all_conditional_entropy) %>% bind_rows()

ggplot(ce, aes(x = X, y = ce, color = as.factor(X))) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_line()

ggplot(ce, aes(x = X, y = ce, color = as.factor(Y))) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_line() + 
  ggtitle("Conditional entropy vs. conditioned random variable")

>>>>

---
title: Positional surprisal (1-3grams) in CHILDES corpora
author: Josef Klafka and Dan Yurovsky 
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
    number_sections: false
    theme: lumen
    toc_float: false
    code_folding: show 
editor_options: 
  chunk_output_type: console
---

@@

Get utterances
```{r get_utterances}
prov_utterances <- get_utterances(corpus = "Providence") #stem or gloss
zhou_utterances <- get_utterances(corpus = "ZhouDinner")
```

```{r}
adult_utterances <- zhou_utterances %>%
  filter(speaker_role == "Target_Child") %>% 
  mutate(length = str_count(gloss, pattern = "[ +]+") + 1) %>%
  mutate(utterance_id = 1:n()) %>%
  unnest_tokens(word, stem) %>%
  group_by(utterance_id) %>%
  mutate(word_order = 1:n())
```

@@

bigram_surprisals_nonnull %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)
```

```{r trigrams}
adult_trigrams <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  group_by(lag_word2, lag_word1, word) %>%
  count() %>%
  filter(!is.na(lag_word1), !is.na(lag_word2)) %>%
  ungroup() %>%
  mutate(tri_joint_p = n / sum(n)) %>%
  select(-n) %>%
  left_join(adult_bigrams, by = c("lag_word2" = "lag_word", "lag_word1" = "word")) %>%
  mutate(tri_cond_p = tri_joint_p / joint_p) %>%
  select(-joint_p, -tri_joint_p, -cond_p) # %>% 
  # left_join(adult_bigrams, by = c("lag_word1" = "lag_word", "word" = "word")) %>% 
  # select(-joint_p)

trigram_surprisals_prep <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  left_join(adult_trigrams) %>% 
  left_join(select(adult_bigrams, -joint_p), by = c("lag_word1" = "lag_word", "word" = "word")) %>% 
  left_join(adult_unigrams)

trigram_surprisals_nonnull <- trigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word2), # check if it's not the first word
                    ifelse(is.na(lag_word1), # check if not the second word
                           -log(p), # trigram if third or beyond
                           -log(cond_p)), # bigram if second word
                    -log(tri_cond_p))) %>% # unigram if first word
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

@@
